---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

Project management and overview documents for the Cycling Infrastructure Prioritisation Toolkit (CyIPT)
Funded by the Department for Transport.

Comments/suggestions welcome in the [issue tracker](https://github.com/cyipt/cyipt/issues).

To get this and associated repos on you computer you can clone them as follows:

```bash
clone git@github.com:cyipt/cyipt
clone git@github.com:cyipt/cyipt-bigdata
```

The main processes undertaken by CyIPT are:

- Download OSM data by Local Authority (by default Bristol). See [download-osm.R](https://github.com/cyipt/cyipt/blob/master/scripts/prep_data/download-osm.R):

```{r}
source("scripts/prep_data/download-osm.R")
```

- Clean OSM data that has been saved to `cyipt-bigdata`:

```{r}
source("scripts/prep_data/clean_osm.R")
```

## Definitions and variable names

Variable names are in lowercase and use_underscores, and should be kept updated in [input-data/variables.csv](https://github.com/cyipt/cyipt/blob/master/input-data/variables.csv):

```{r, message=FALSE}
variables = readr::read_csv("input-data/variables.csv")
knitr::kable(variables)
```

